
EDA of Prosper Loan Data By Ramamohan Soma
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

#if you need to install the packages please uncomment the below.

#install.packages('ggplot2')
#install.packages('dplyr') 
#install.packages('gridExtra')
#install.packages('alr3')
#install.packages('reshape2')
#install.packages('scales')
#install.packages('GGally')
#install.packages('memisc')
#install.packages('RColorBrewer')
#install.packages('plyr')

suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(gridExtra))
suppressMessages(library(alr3))
suppressMessages(library(reshape2))
suppressMessages(library(scales))
suppressMessages(library(GGally))
suppressMessages(library(memisc))
suppressMessages(library(RColorBrewer))
suppressMessages(library(plyr))

knitr::opts_chunk$set(warning=FALSE, message=FALSE)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data

setwd('C:/nanodegree/DA/EDA/P3Project')

LoanDataAll <- read.csv('prosperLoanData.csv',
                        header = TRUE, 
                        check.names = FALSE)

LoanData <- LoanDataAll %>%
  dplyr::select(Term,LoanStatus,BorrowerRate,BorrowerState,Occupation,
                EmploymentStatus,IsBorrowerHomeowner,CreditScoreRangeLower,
                CreditScoreRangeUpper,CurrentDelinquencies,AmountDelinquent,
                IncomeRange,LoanCurrentDaysDelinquent,LoanOriginalAmount,
                LoanOriginationQuarter,MonthlyLoanPayment,
                LP_CustomerPayments,LP_CustomerPrincipalPayments,
                LP_InterestandFees,LP_NetPrincipalLoss)

LoanData$Year <- sub("[[:alnum:]][[:digit:]] ",
                     "",
                     LoanData$LoanOriginationQuarter)


dfTemp <- subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
                   !is.na(CurrentDelinquencies))

# Data with deliquencies between 1-5
dfTemp1 <- subset(LoanData,LoanData$CurrentDelinquencies > 0 
                  & LoanData$CurrentDelinquencies < 5 &
                   !is.na(CurrentDelinquencies))
# Data with deliquencies between 5-50
dfTemp2 <- subset(LoanData,LoanData$CurrentDelinquencies > 4 
                  & LoanData$CurrentDelinquencies < 51 &
                   !is.na(CurrentDelinquencies))

```

I have chosen the Propser Loan data for this project to explore. There are 113,937 loans (rows) with 81 variables. Below are the names of all the variables.


```{r echo=FALSE, Loan_info}

names(LoanDataAll)

```

Below are the variables from the data I have chosen to analyse and more information on those variables.

```{r echo=FALSE, Loan_info1}

names(LoanData)
str(LoanData)

levels(LoanData$LoanStatus)
levels(LoanData$BorrowerState)
levels(LoanData$EmploymentStatus)
levels(LoanData$IsBorrowerHomeowner)
levels(LoanData$IncomeRange)
levels(LoanData$LoanOriginationQuarter)

```


# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots1}

# Histogrman for LoanOriginalAmount

ggplot(LoanData, aes(x = LoanOriginalAmount)) +
  geom_histogram(binwidth = 400, color = 'black',fill = '#099DD9') +
  scale_x_continuous(breaks = seq(0,40000, 5000),
                     limits = c(0,40000), labels = dollar) +
  xlab('Original Loan Amount (LoanOriginalAmount)') +
  ylab('Number of Cutomers Borrowed (count)') +
  ggtitle('LoanAmounts\n')

```

From the above plots it is evident that for most of the Loan Amonts the number of borrowers are below 7500. Also there is spike in users for the loan amounts around 4,000, 10,000, and 15,000 dollars. Below is the summary of LoanOriginalAmount. Mean in 8337 and Median is 6500.

```{r echo=FALSE, Univariate_Plots1-1}

#Summary and Range of Loan Amount
summary(LoanData$LoanOriginalAmount)
range(LoanData$LoanOriginalAmount)

```

Let's look at the terms of the loans that people were borrowing.

```{r echo=FALSE, Univariate_Plots2}

# Histogrman for Loan Term
# As per coach comment changed the y-axis to log scale.
# Also made a anothe plot with discrete x-axis.

ggplot(LoanData, aes(x = Term)) +
  geom_histogram(binwidth = 5, fill = '#099DD9',
                 position='identity',drop=TRUE) +
  scale_x_continuous(breaks = seq(0,72, 12),limits = c(0,72)) +
  scale_y_log10() +
  xlab('Loan Term') +
  ylab('Number of Loans (log10)') +
  ggtitle('Loan Terms\n')


# Number of Loans for each Term.
table(LoanDataAll$Term)

```

I see that there are 3 types of Terms (12m, 36m, 60m) and 36months loans are famous(approximately 80% of the loans).

```{r echo=FALSE, Univariate_Plots3}

#ggplot(LoanData, aes(x = BorrowerRate)) +
#  geom_histogram(binwidth = .01, color = 'black', fill = '#099DD9') +
#  scale_x_continuous(breaks = seq(0.04,0.5, 0.01),limits = c(0.04,0.5)) +
#  scale_y_continuous(breaks = seq(0,7000, 200),limits = c(0,7000)) +
#  ggtitle('Loan Rates')

# Histogrman for Borrower Loan Rates

ggplot(LoanData, aes(x = BorrowerRate)) +
  geom_histogram(binwidth = .005, color = 'black', 
                 fill = '#099DD9') +
  scale_x_continuous(breaks = seq(0.04,0.5, 0.01),
                     limits = c(0.04,0.5)) +
  scale_y_continuous(breaks = seq(0,7000, 500),
                     limits = c(0,7000)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab('Borrower Rate') +
  ylab('Number of Loans') +
  ggtitle('Loan Rates\n')

```

From the above graph on Borrower Rates, Very few loans (4 or 5) are there with 0.4 and above. That's why not visible in the plot. Range is from 0 to 4.975. Median Borrower rate is 0.184. There is spike of count for Borrower rate 0.3177. I'll investigate more on this in BiVariate plots section.

Summary and range of Borrower Rate.

```{r echo=FALSE, Univariate_Plots3-1}

#Summary of BorrowerRate and range.
summary(LoanData$BorrowerRate)
range(LoanData$BorrowerRate)


```

Below is the histogram of CreditScoreRangeUpper. Lets see how the number of loans varies based on upper range credit score.

```{r echo=FALSE, Univariate_Plots4}

#Histogram of CreditScoreRangeUpper.

ggplot(aes(x = CreditScoreRangeUpper), data = LoanData) +
  geom_histogram(binwidth = 10, color = 'black', 
                 fill = '#099DD9') +
  scale_x_continuous(breaks = seq(400, 900, 50),
                     limits=c(400,900)) +
  scale_y_continuous(breaks = seq(0, 15000, 5000)) +
  xlab('Upper Range Credit Score (CreditScoreRangeUpper)') +
  ylab('Number of Cutomers Borrowed (count)') +
  ggtitle('Borrowers with Upper Range Credit Score\n')


```

Borrowers with upper range credit score 650-800 have most of the loans.Borowers with upper range credit score of 679,699 have maximum number of loans. Median is 699. Below is the summary.

```{r echo=FALSE, Univariate_Plots4-1}

# Summary of CreditScoreRangeUpper
summary(LoanData$CreditScoreRangeUpper)
table(LoanData$CreditScoreRangeUpper)


```

Let's look at the status the of the Loans borrowed.

```{r echo=FALSE, Univariate_Plots5}

# Histogram of Loans Status
# As per coach comment changed the y-axis to log scale.

ggplot(LoanData, aes(x = LoanStatus)) +
  geom_histogram(binwidth = 100, fill = '#099DD9') +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab('Loan Status') +
  ylab('Number of Loans (log10)') +
  ggtitle('Status of the Loans Borrowed\n')


```

In the above plot most loans are either completed or current or chargedoff(~10%). There are approximately 5% defaulted loans.Below is info on each loan status.


```{r echo=FALSE, Univariate_Plots5-1}

# Summary of each Loan Status
summary(LoanData$LoanStatus)

```

I would like to see howmany loans have been Delinquent. Below is the plot for current Delinquencies of the loans.

```{r echo=FALSE, Univariate_Plots6}

# Frequency Polygon for CurrentDeliquencies
# As per coach comment changed the y-axis to log scale.

ggplot(subset(LoanData, 
              LoanData$CurrentDelinquencies > 0 & 
                !is.na(CurrentDelinquencies)),
       aes(x = CurrentDelinquencies)) +
  geom_freqpoly(binwidth = 1) +
  scale_x_continuous(breaks = seq(1, 83, 3)) +
  scale_y_log10() +
  xlab('Current Deliquencies') +
  ylab('Number of Loans(log10)') +
  ggtitle('Borrowers Deliquencies\n')

```


I see that ~12000 customers with 1 Deliquency. Interestingly numberof cusotmers decreased with the increase in Deliquency count. But still total number of Deliquencies is 67,044 which is high. Range is from 1 to 83. Not sure why so many Deliquencies even though loans were issued based on credit score. There are spikes for some deliquencies after 28.

Below is the Summary of CurrentDelinquencies with value > 0.

```{r echo=FALSE, Univariate_Plots6-1}

#summary(LoanData$CurrentDelinquencies)
#table(LoanData$CurrentDelinquencies)
#sum(LoanData$CurrentDelinquencies,na.rm=TRUE)

#Information on deliquencies whose value is greater than 0.

summary(subset(LoanData,LoanData$CurrentDelinquencies>0 & 
                 !is.na(CurrentDelinquencies))$CurrentDelinquencies)

table(subset(LoanData,LoanData$CurrentDelinquencies>0 & 
               !is.na(CurrentDelinquencies))$CurrentDelinquencies)

sum(subset(LoanData,LoanData$CurrentDelinquencies>0 & 
             !is.na(CurrentDelinquencies))$CurrentDelinquencies)

```

I want to look at the principal Loss of the loans to the investors. Below is the plot using the variable LP_NetPrincipalLoss which means the principal remains uncollected after any recoveries.

```{r echo=FALSE, Univariate_Plots7}

#ggplot(LoanData, aes(x = LP_CustomerPayments)) +
#  geom_histogram(binwidth = 500,color = 'black',fill = '#099DD9') +
#  scale_x_continuous(breaks = seq(0, 35000, 1000)) +

# Histogram of Loan Net Principal Loss whose value is > 0.

ggplot(subset(LoanData, 
                LoanData$LP_NetPrincipalLoss > 0),
       aes(x = LP_NetPrincipalLoss)) +
  geom_histogram(binwidth = 100,color = 'black',fill = '#099DD9') +
  scale_x_continuous(labels = dollar) +
  xlab('Net Principal Loss (LP_NetPrincipalLoss)') +
  ylab('Number of Loans with Principal Loss (count)') +
  ggtitle('Loss of princial Amount on the Loans Issued\n')


```

Above plot is made with the values of 'LP_NetPrincipalLoss' > 0. Most of the Principal Loss is under $5000. There are loans with Principal Loss between $5000 and $15000. NUmber of Loans with Principal Loss between $1500 to $25000 is much less compared to others. 

Summary of 'LP_NetPrincipalLoss' > 0:

```{r echo=FALSE, Univariate_Plots7-1}

#Information on LP_NetPrincipalLos whose value is greater than 0.

summary(subset(LoanData,
               LoanData$LP_NetPrincipalLoss>0)$LP_NetPrincipalLoss)

```


I have observed that there are some loans with 'LP_NetPrincipalLoss' < 0 which means it's a Principal Profit exculding Interest and Fees. Below is the plot. 


```{r echo=FALSE, Univariate_Plots8}

# Histogram of Loan Net Principal Loss whose value is < 0 which means
# it's a profit.
# As per coach comment changes y-axis to log scale and introduced
# breaks for x-axis.

ggplot(subset(LoanData, 
              LoanData$LP_NetPrincipalLoss < 0),
       aes(x = LP_NetPrincipalLoss)) +
  geom_histogram(binwidth = 100,color = 'black',
                 fill = 'orange',position = "identity") +
  scale_x_continuous(breaks=seq(-1000,0,100), 
                     limits=c(-1000,0), labels = dollar) +
  scale_y_log10() +
  xlab('Net Principal Gain (LP_NetPrincipalLoss < 0)') +
  ylab('Number of Loans with Principal Gain (count)') +
  ggtitle('Gain of princial Amount on the Loans Issued\n')


```

Summary of 'LP_NetPrincipalLoss' < 0:

```{r echo=FALSE, Univariate_Plots8-1}

#Information on LP_NetPrincipalLos whose value is less than 0.
summary(subset(LoanData,
               LoanData$LP_NetPrincipalLoss<0)$LP_NetPrincipalLoss)
```

The Median Gain in Net Princial is $23.82. There are loans with gains above $500 but very less. They are shown below 1 in the plot as y-axis is a log scale. We can drop those by using drop="TRUE" as parameter in geom_histogram but I wanted to keep those so that reviewer would know there are entries with Gain >$500.


Now let's look at the Numer of loans borrowed each Year.

```{r echo=FALSE, Univariate_Plots9}

# Bar plot of Loan Year

ggplot(aes(x = Year), data = LoanData) +
  geom_histogram(aes(fill = ..count..)) +
  scale_y_continuous(breaks = seq(0, 35000, 5000)) +
  xlab('Loan Year') +
  ylab('Number of Cutomers Borrowed (count)') +
  ggtitle('Yearly Loans\n')

```


I have extracted the Year from the LoanOriginationQuarter and plotted the above histogram. Year 2013 have the maximum number of loans and 2009 have the lowest number of Loans. I need to investigate the reason for having so less number in 2009. Below is the number of Loans per year. I also see that in 2005 there were 22 loans. May be we can ignore that.


```{r echo=FALSE, Univariate_Plots9-1}
# Summary of Loans per year.

table(LoanData$Year)

```


# Univariate Analysis

### What is the structure of your dataset?

There are 113,937 loans in the Data set and 81 features. I have selected the below 17 features.

Term, LoanStatus, BorrowerState, EmploymentStatus, IsBorrowerHomeowner, LoanOriginationQuarter,
BorrowerRate,CurrentDelinquencies,AmountDelinquent,IncomeRange,LoanCurrentDaysDelinquent,LoanOriginalAmount, MonthlyLoanPayment, LP_CustomerPayments, LP_CustomerPrincipalPayments, LP_InterestandFees, LP_NetPrincipalLoss

Minimum LoanAmount Borrowed is $1,000 and Maximum is $35,000. There are 3 Loan Terms (12 months, 36 months, 60 months). 36 Months loans are popular. Median Borrower rate is 18%(0.18). Employed and FullTime are the main Borrowers category who most took loans. Most of the Loans are either Current or Completed or ChargedOff. There are approximately 12,000 borrowers with 1 CurrentDeliquency and there total 67,044 current deliquencies which is high. There are 697 Current Deliquencies with 'NA'. The Median LP_NetPrincipalLoss(The principal that remains uncollected after any recoveries) is $3265. Income Range of the most Borrowers is from $25,000 to $75,000.Year 2013 have the maximum number of loans and 2009 have the lowest number of Loans.


### What is/are the main feature(s) of interest in your dataset?

I am interested in many features from the data set but I have isolated to investigate the below features which tells about the Loan Amounts borrowed, Loan rates, Deliquencies etc.

BorrowerRate,CurrentDelinquencies,AmountDelinquent,IncomeRange,LoanCurrentDaysDelinquent,LoanOriginalAmount, MonthlyLoanPayment, Term, LoanStatus LP_NetPrincipalLoss, CreditScoreRangeUpper.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

I think Year of the loan would be helpful to see how many Current Deliquencies per year and how many Net Principal losses per year.

### Did you create any new variables from existing variables in the dataset?

Yes! I have created a new variable 'year' in which the loan was borrowed.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

One thing I have observed is the number of lOans have decreased significantly in 2009. I need to investigate more on this and find the reason. I have extracted the year from LoanOriginationQuarter. Also for LP_NetPrincipalLoss, there are values which are > 0 (means loss) and < 0 (means gain). so in future plots I am planning to use LP_NetPrincipalLoss values with > 0.

# Bivariate Plots Section

First I want to explore the relationship between the LoanOriginalAmount and the CurrentDeliquencies.


```{r echo=FALSE, Bivariate_Plots1}

# Scatter plot of LoanOriginalAmount vs CurrentDelinquencies. Limit the
# x scale to 40,000 with breaks of 5000.

ggplot(aes(x = LoanOriginalAmount, y = CurrentDelinquencies), 
       data = subset(LoanData,
                     LoanData$CurrentDelinquencies > 0 & 
                       !is.na(CurrentDelinquencies))) +
  geom_jitter(alpha=1/10,
              position=position_jitter(h=0),
              color='red') +
  scale_x_continuous(breaks = seq(0,40000, 5000),
                     limits = c(0,40000), labels = dollar) +
  ggtitle("Delinquencies by Loan Amount\n")

# Pearson correlation test on LoanOrignalAmount and CurrentDelinquencies.

cor.test(LoanData$LoanOriginalAmount,
         LoanData$CurrentDelinquencies,
         method='pearson')

```

From the plot it seems LoanAmounts under $10,000 has many deliquencies. We can see straight lines in the plot for $10,000, $15,000, $20,000 and $25,000 which indicates there are deliquencies even for big amounts. Don't see a good correlation between these two paramters. Correlation value is -0.1721394.

Let's look at the CurrentDeliquencies versus Term Relationship to see which Term loans have more delinquencies.


```{r echo=FALSE, Bivariate_Plots2}

# Scatter Plot between Term and CurrentDelinquencies with 
# values >0 and excluding NA's.
# As per Coach comment converted the plot to Box Plot.


#ggplot(aes(x = Term, y = CurrentDelinquencies), 
#       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_point() +
#  scale_y_continuous(breaks = seq(1, 83, 3)) +
#  ggtitle("Delinquencies by Term\n")

# As per review comment I changes the y-axis scale 
# range to 5-20 to remove outliers. Before it was 1-20. 
# There are number of deliquencies with 1.

bp2 <- ggplot(aes(x = factor(Term), y = CurrentDelinquencies),
              data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
                       !is.na(CurrentDelinquencies))) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(5, 20, 5),limits=c(5,20)) +
  xlab("Term") +
  ylab("CurrentDelinquencies") +
  ggtitle("Delinquencies by Term\n")

suppressWarnings(print(bp2))

cor.test(dfTemp$Term,dfTemp$CurrentDelinquencies,method='pearson')


```

In the above plot I have limited the y-axis scale to 5-20 to get a better picture of the Box Plot and to remove the entries with deliquencies under 5. Plot depicts that Delinquencies are more for 36 months Term Loans than other Terms. Correlation Value is -0.0838 and the value without zero's and 'NA's is -0.0996. Below is the number of deliquencies for each Term.

```{r echo=FALSE, Bivariate_Plots2-1}

# Number of Delinquencies per Term.
by(dfTemp$CurrentDelinquencies,dfTemp$Term,sum)

```

Now I would like see how the number of Deliquencies varies by Income Range.

```{r echo=FALSE, Bivariate_Plots3}

# Box plot of IncomeRange vs CurrentDelinquencies > 0 with fill as
# IncomeRange. Limit the deliquencies to 83 with a break of 2.
# As per coach comments I have modified to use scale_fill_brewer()
# on Incomerange. Also converted y-axis to log10 so that
# box plots are clear.

#ggplot(aes(x = IncomeRange,
#           y = CurrentDelinquencies,
#           fill=IncomeRange), 
#       data = subset(LoanData,
#                     LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_boxplot() +
#  theme(axis.text.x = element_text(angle = 90)) +
#  scale_y_continuous(limits = c(1,83),breaks = seq(1,83,2)) +
#  ylab('CurrentDelinquencies') +
#  xlab('Income Range of Borrowers(IncomeRange)') +
#  ggtitle("Number of Current Delinquencies by Borrower Income Range\n")

#ggplot(aes(x = IncomeRange,
#           y = CurrentDelinquencies,
#           fill=IncomeRange), 
#       data = subset(LoanData,
#                     LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_boxplot() +
#  scale_fill_brewer() +
#  theme(axis.text.x = element_text(angle = 90)) +
#  scale_y_log10() +
#  ylab('CurrentDelinquencies') +
#  xlab('Income Range of Borrowers(IncomeRange)') +
#  ggtitle("Number of Current Delinquencies by Borrower Income Range\n")

# As per review comment I changes the y-axis scale 
# range to 5-50 to remove data below 5. 
# There are number of deliquencies with 1.

ggplot(aes(x = IncomeRange,
           y = CurrentDelinquencies,
           fill=IncomeRange), 
       data = dfTemp2) +
  geom_boxplot() +
  scale_fill_brewer() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(breaks = seq(5, 50, 5),limits=c(5,50)) +
  ylab('CurrentDelinquencies') +
  xlab('Income Range of Borrowers(IncomeRange)') +
  ggtitle("Number of Current Delinquencies by Borrower Income Range\n")

```

Below is the details of CurrentDeliquencies(range 5-50) for each Income Range.

```{r echo=FALSE, Bivariate_Plots3-1}

# Summary of CurrentDelinquencies based on IncomeRange
by(dfTemp2$CurrentDelinquencies,dfTemp2$IncomeRange,summary)

```

Here I have taken the subset of data with number of delinquencies range 5-50 so that entries with 1 are eliminated.Median and Min Value of CurrentDeliquencies for Income Ranges $1-24,999, $25,000-49,999,$50,000-74,999, $75,000-99,999, $100,000+ and Not Employed are same. Median value for Borrowers with Income $0 is 8.Surprisingly for Borrowers whose Income Range is Not Displayed is 8 and Maximum is 41.   Maximum for below Mid Range Income($25,000-49) is 50.


```{r echo=FALSE, Bivariate_Plots4}

# Box plot of EmploymentStatus vs CurrentDelinquencies > 0 with fill as
# EmploymentStatus. Limit the deliquencies to 83 with a break of 2.

# As per review comment I changes the y-axis scale 
# range to 5-50 to remove data below 5. 
# There are number of deliquencies with 1.

ggplot(aes(x = EmploymentStatus,
           y = CurrentDelinquencies,
           fill=EmploymentStatus), 
       data = dfTemp2) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(breaks = seq(5, 50, 5),limits=c(5,50)) +
  ylab('CurrentDelinquencies') +
  xlab('Borrowers Employment Status(EmploymentStatus)') +
  ggtitle("Current Delinquencies by Borrower's Employment Status\n")



```

Below is the details of CurrentDelinquencies(range 5-50) for each Employment Status.

```{r echo=FALSE, Bivariate_Plots4-1}

# Summary of CurrentDelinquencies based on EmploymentStatus
by(dfTemp2$CurrentDelinquencies,dfTemp2$EmploymentStatus,summary)

```

Here I have taken the subset of data with number of delinquencies range 5-50 so that entries with 1 are eliminated.
Full Time Employed Borrowers have a median of 7 and Maximum of 50. There is a group of Borrowers whose employment status is 'Not Available'. Median for them is 8 and Maximum is 41 which is pretty high. There are many Borrowers whose Employment Status is blank and they have a median of 8 and Maximum of 39.

Now Lets see how the number of CurrentDelinquencies varies by Year.

```{r echo=FALSE, Bivariate_Plots5}

# Box plot of CurrentDelinquencies > 0 per year with fill as
# Year. 
# As per review comment I changes the y-axis scale 
# range to 5-50 to remove data below 5. 
# There are number of deliquencies with 1.

ggplot(aes(x = Year, y = CurrentDelinquencies),
       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
                       !is.na(CurrentDelinquencies))) +
  geom_boxplot(aes(fill = Year)) +
  scale_y_continuous(breaks = seq(5, 50, 5),limits=c(5,50)) +
  scale_fill_brewer(palette='OrRd') +
  ylab('CurrentDelinquencies') +
  xlab('Year') +
  ggtitle('CurrentDelinquencies by Year\n')

```

Below is the number of CurrentDelinquencies(range 5-50) per year.

```{r echo=FALSE, Bivariate_Plots5-1}

# Sum of Deliquencies per year
by(dfTemp2$CurrentDelinquencies,dfTemp2$Year,sum)

```

Here I have taken the subset of data with number of delinquencies range 5-50. Number of CurrentDelinquencies were high in 2006 and 2007. There's a decrease in the count significantly in 2009 but again picked up in later years. 

I want to see the relationship between CurrentDelinquencies and CreditScoreRangeUpper. Below is the plot.

```{r echo=FALSE, Bivariate_Plots5-2}

#ggplot(aes(x = factor(CreditScoreRangeUpper), 
#           y = CurrentDelinquencies),
#       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_boxplot(aes(fill = CreditScoreRangeUpper),width=1.0) +
#  theme_bw() +
#  scale_y_continuous(breaks=seq(0,85,5),limits=c(0,85)) +
#  theme(axis.text.x = element_text(angle = 90)) +
#  ylab('CurrentDelinquencies') +
#  xlab('CreditScoreRangeUpper') +
#  ggtitle('CurrentDelinquencies by CreditScoreRangeUpper\n')

# As per review comment I changes the y-axis scale 
# range to 5-50 to remove data below 5. 
# There are number of deliquencies with 1.

ggplot(aes(x = factor(CreditScoreRangeUpper), 
           y = CurrentDelinquencies),
       data = dfTemp2) +
  geom_boxplot(aes(fill = CreditScoreRangeUpper),width=1.0) +
  theme_bw() +
  scale_y_continuous(breaks = seq(5, 50, 5),limits=c(5,50)) +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab('CurrentDelinquencies') +
  xlab('CreditScoreRangeUpper') +
  ggtitle('CurrentDelinquencies by CreditScoreRangeUpper\n')



cor.test(dfTemp$CreditScoreRangeUpper,dfTemp$CurrentDelinquencies,method='pearson')

# Sum of Deliquencies per year
by(dfTemp2$CurrentDelinquencies,dfTemp2$CreditScoreRangeUpper,summary)

by(dfTemp2$CurrentDelinquencies,dfTemp2$CreditScoreRangeUpper,sum)


```

Here also I have taken the subset of data with number of delinquencies range 5-50. In the above plot, the  median of delinquencies is high at credit score of 459, 479, 499, 519 and dereased by increase in credit score upto 699 . Borrowers with credit score of 539 has highest sum of deliquencies. The number of deliquencies decreased from 759.

Lets look at the relationship between LoanOriginalAmount and MonthlyLoanPayment.

```{r echo=FALSE, Bivariate_Plots6}

# Scatter Plot between LoanOriginalAmount and MonthlyLoanPayment. 
# due to overplotting used the 'alpha' parameter in the gemo_point 
# and introduced some jitter.

ggplot(aes(x = LoanOriginalAmount, y = MonthlyLoanPayment), 
       data = LoanData) +
  geom_point(alpha=1/4,size = 1, position = 'jitter') +
  geom_smooth(method='lm',color='red') +
  xlab('LoanOriginalAmount') +
  ylab('MonthlyLoanPayment') +
  ggtitle('Loan Amount by Monthly Payment\n')

#Pearson Correlation test on LoanOriginalAmount and MonthlyLoanPayment
cor.test(LoanData$LoanOriginalAmount,
         LoanData$MonthlyLoanPayment,
         method='pearson')

```

There is a strong correlation between LoanOriginalAmount and MonthlyLoanPayment and the value is 0.932(rounded). It might be that as the Monthy Payment increases as Loan Amount increases. I see a line where there is steep increse in monthly payments as Loan Amount increases. It could be that BorrowerRate might also affect the relationship.

```{r echo=FALSE, Bivariate_Plots7}

# Box plot of IsBorrowerHomeowner vs LoanOriginalAmount with fill
# as IsBorrowerHomeowner. Limit the scale for LoanOriginalAmount to
# $35000 with breaks of $5000

# As per review comment redcued the scale for LoanOriginalAmount to
# $20,000 with breaks of $5000.

ggplot(aes(x = IsBorrowerHomeowner,
           y = LoanOriginalAmount,
           fill=IsBorrowerHomeowner ), 
       data = LoanData) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0,20000),
                     breaks = seq(0,20000,5000),
                     labels=dollar) +
  ylab('Original Loan Amount (LoanOriginalAmount)') +
  xlab('Is Borrower Homeowner') +
  ggtitle("LoanAmount by IsBorrowerHomeowner\n")

# Summary of Loan Amounts and Number of Loans
# for Homeowners and Non Homeowners.

by(LoanData$LoanOriginalAmount,
   LoanData$IsBorrowerHomeowner,
   summary)
summary(LoanData$IsBorrowerHomeowner)

```

In the above box plot, y-axis is limited to $35000 with breaks of $5000 to take a closer look.Then the statistics(Median,1st Quartuile,3rd Quartile) from 'by' command were matching the Box plots for Homeowners and Non Homeowners. Median, Mean and 3rd Quartile values are more for Homeowners which they are borrowing more money than non Homeowners. A little bit more than half of the borrowers are homeowners.

Below is the histogram of LoanOriginalAmount by EmploymentStatus.

```{r echo=FALSE, Bivariate_Plots8}

#Histogram of Loan Amounts faceted by Borrower's Employment Status.
# As per review comment changed the y-axis to log scale.

ggplot(aes(x = LoanOriginalAmount), data = LoanData) +
  geom_histogram(binwidth = 500, color = 'black', 
                 fill = '#099DD9',drop="TRUE") + 
  scale_x_continuous(breaks = seq(0, 35000, 5000), labels = dollar) +
  coord_cartesian(c(0,35000)) +
  scale_y_log10() +
  facet_grid(EmploymentStatus~.,scales = "free_y") +
  theme(strip.text.y = element_text(size=8, face="bold")) +
  xlab('Original Loan Amount (LoanOriginalAmount)') +
  ylab('Number of Cutomers Borrowed (count)') +
  ggtitle('LoanAmounts by EmploymentStatus\n')

#max(LoanData$LoanOriginalAmount)
```

In the above plot I have y axis scales free and log10 so that bars appear properly. Plot depicts that Employed and FullTime borrowers are more than others. Some borrowr types are 'Not available' and 'Other'. Self Employed borrowed less and PartTime/Ritired users did not borrow at all.In all the categories I see that there are many borrowers for the loan amounts under $10,000. For Bigger Loan amounts like $20,000 and $25,000 I see Employed, Fulltime and 'Not available' have some. Borrowers for Loan Amounts above $25,000 are not seen in the graph as there might be very few less than the minimum scale in the plot. 

Now let's look at the Income Range of the People who has borrowed loans and also if they are Homeowners or not.


```{r echo=FALSE, Bivariate_Plots9}

# Bar plot of Income Range, fill with IsBorrowerHomeowner
# As per review comments added postion='dodge'

ggplot(LoanData, mapping = aes(x = IncomeRange)) +
  layer(geom = "bar", 
        mapping = aes(fill = IsBorrowerHomeowner),
        position="dodge") +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab('Income Range') +
  ylab('Number of Loans(count)') +
  ggtitle('Income Range of the Borrowers\n')


```

Income Range of the most Borrowers is in the middle ($25k to $75k) and many of them are not homeowners. Aslo with mostly 36 month loans as per the data.Surprisingly there are loans with zero Income Range of. Does that mean Borowers without income can get loans?

Below is the summary of Income range of the borrowers.

```{r echo=FALSE, Bivariate_Plots9-1}

# Summary of Income Range.

summary(LoanData$IncomeRange)

```

In the UniVariate Plot on BorrowerRate, I saw a spike for the rate of 0.3177. I want to explore more on this. Below is the plot on BorrowerRate based on EmploymentStatus.

```{r echo=FALSE, Bivariate_Plots10}

# Bar plot with x = BorrowerRate and fill as  EmploymentStatus. 
# Adjusted the x-axis 0.3 to 0.37 and y-axis 0 to 4000 to closely look 
# at the spike for rate 0.3177.


#bp10 <- ggplot(LoanData, mapping = aes(x = BorrowerRate)) +
#  layer(geom = "bar", 
#        mapping = aes(fill = EmploymentStatus),
#        stat = "bin",
#        position="stack",
#        stat_params = list(binwidth = .001)) +
#  scale_x_continuous(breaks = seq(0.3,0.37, 0.01),
#                     limits = c(0.3,0.37)) +
#  scale_y_continuous(breaks = seq(0,4000, 500),
#                     limits = c(0,4000)) +
#  xlab('Borrower Rate') +
# ylab('Number of Loans(count)') +
#  ggtitle('Number of Loans by Borrower Rate\n')

# As per Review comments added position=dodge and
# adjusted the x lim and y lim

bp10 <- ggplot(LoanData, 
               mapping = aes(x = BorrowerRate,
                                       fill = EmploymentStatus)) +
  layer(geom = "bar", 
        stat = "bin",
        position="dodge",
        stat_params = list(binwidth = 0.0001)) +
  scale_fill_brewer(palette="OrRd") +
  scale_x_continuous(breaks = seq(0.317,0.318, 0.0001),
                       limits = c(0.317,0.318)) +
  scale_y_continuous(breaks = seq(0,4000, 500),
                     limits = c(0,4000)) +
  xlab('Borrower Rate') +
  ylab('Number of Loans(count)') +
  ggtitle('Number of Loans by Borrower Rate\n')

suppressWarnings(print(bp10))

```

I see that Many Employed Borrowers got this rate. I see some 'other' and 'SelfEmployed' borrowers also have this rate.

Let's see which income range borrowers has this rate. Below is the plot on BorrowerRate based on IncomeRange.

```{r  echo=FALSE, Bivariate_Plots11}

# Bar plot with x = BorrowerRate and fill as  IncomeRange. 
# Adjusted the x-axis 0.3 to 0.4 to closely look at the spike 
# for rate 0.3177.
# As per coach comments,i have reduced the x and y-axis scale
# and used scale_fill_brewer on IncomeRange for a clear view.


#ggplot(LoanData, mapping = aes(x = BorrowerRate,fill = IncomeRange)) +
#  layer(geom = "bar", 
#        stat = "bin",
#        position="stack",
#        stat_params = list(binwidth = 0.001)) +
#  scale_x_continuous(breaks = seq(0.3,0.4, 0.01),
#                     limits = c(0.3,0.4)) +
#  scale_y_continuous(breaks = seq(0,7000, 1000),
#                     limits = c(0,7000)) +
#  xlab('Borrower Rate') +
#  ylab('Number of Loans(count)') +
#  ggtitle('Number of Loans by Borrower Rate\n')

# As per Review comments added position=dodge and
# adjusted the y lim
  
ggplot(LoanData, mapping = aes(x = BorrowerRate,
                               fill = IncomeRange)) +
  layer(geom = "bar", 
        stat = "bin",
        position="dodge",
        stat_params = list(binwidth = 0.001)) +
  scale_fill_brewer(palette="OrRd") +
  scale_x_continuous(breaks = seq(0.31,0.32, 0.001),
                       limits = c(0.31,0.32)) +
  scale_y_continuous(breaks = seq(0,1500, 300),
                     limits = c(0,1500)) +
  xlab('Borrower Rate') +
  ylab('Number of Loans(count)') +
  ggtitle('Number of Loans by Borrower Rate\n')


```

I see that except Income range of $0 and 'Not Displayed' all other Income Ranges borrowers have the rate 0.3177.

Let's see the behaviour of LP_NetPrincipalLoss(The principal that remains uncollected after any recoveries.) against the LoanOriginalAmount.

```{r echo=FALSE, Bivariate_Plots12}

# Scatter plot of LP_NetPrincipalLoss vs LoanOriginalAmount.
# For overplotting used alpha=1/10 in geom_point().

LoanDataPL <- subset(LoanData,LoanData$LP_NetPrincipalLoss>0)

ggplot(aes(x = LP_NetPrincipalLoss, y = LoanOriginalAmount), 
       data = LoanDataPL) +
  geom_point(alpha=1/10,size = 3, position = 'jitter', color="blue") +
  ggtitle('LP_NetPrincipalLoss vs LoanOriginalAmount\n') +
  xlab('LP_NetPrincipalLoss') +
  ylab('LoanOriginalAmount') +
  ggtitle('LP_NetPrincipalLoss vs LoanOriginalAmount\n')

# Scatter plot of LP_NetPrincipalLoss vs log10(LoanOriginalAmount).
# For overplotting used alpha=1/10 in geom_point().

ggplot(aes(x = LP_NetPrincipalLoss, y = LoanOriginalAmount), 
       data = LoanDataPL) +
  geom_point(alpha=1/10,size = 3, 
             position = 'jitter', color="blue") +
  coord_trans(y = 'log10') +
  xlab('LP_NetPrincipalLoss') +
  ylab('log10(LoanOriginalAmount)') +
  ggtitle('LP_NetPrincipalLoss vs log10(LoanOriginalAmount)\n')

# Pearson Correlation test on LoanOriginalAmount and LP_NetPrincipalLoss
cor.test(LoanDataPL$LoanOriginalAmount,
         LoanDataPL$LP_NetPrincipalLoss,
         method='pearson')


```

I see a good correlation value of 0.889 between LP_NetPrincipalLoss and LoanOriginalAmount. LP_NetPrincipalLoss is high for Larger Loans.

Let's see how the LP_NetPrincipalLoss varies each year.

```{r echo=FALSE, Bivariate_Plots13}

# Box plot of year vs LP_NetPrincipalLoss.

ggplot(aes(x = factor(Year), y = LP_NetPrincipalLoss), 
       data = LoanDataPL) +
  geom_boxplot(aes(fill = Year)) +
  scale_fill_brewer(palette='OrRd') +
  xlab('Year') +
  ylab('LP_NetPrincipalLoss') +
  ggtitle('year vs LP_NetPrincipalLoss\n')

```

From the above plot I see that the median Net Principal Loss was high initially  from 2006 to 2008 and it dropped in 2009. Then it gradually increased every year. Below is the summary of Net Principal Loss each year.

```{r echo=FALSE, Bivariate_Plots13-1}

by(LoanDataPL$LP_NetPrincipalLoss,
   LoanDataPL$Year,
   summary)


```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

- Relationship between LoanOrinalAmount and CurrentDeliquencies is not linear. Most of the current deliquencies are for Loan Amounts under $10,000. Ofcourse deliquencies do exists for higher Loan amounts from $15,000 to $35,000.
- Most of the Loans issued are 36 months Term and even the number of Deliquencies is high for these loans.
- Mid income range borowers have most deliquencies but surprisingly borrowers whose income range is not displayed also have a maximum of 82 deliquencies.
- Similary borrowers whose Employment Status is not displayed have signigicant deliquencies.
- There is a strong correlation between Monthly Loan Payment and the Loan Amount which should be and borrowers who are home owners have more Loans.
- Relationship between LoanOriginalAmount and LP_NetPrincipalLoss is also very strong.
- Net Principal Loss was high initially from 2006 to 2008, it dropped in 2009 and increased from 2010 to 2013.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Yes! I did observe some interesting relationship when I plotted the box plot between Year of the Loan originated vs CurrentDeliquencies, NetPrincipalLoss. There was sudden drop of Delinquencies and Net Principal Loss in the year 2009. On more investigation I found that in 2009 there was a relaunch of prosper and implemented stricter guide lines which caused the default rate significantly reduced.

### What was the strongest relationship you found?

The stronges relationships I found was between LoanoriginalAmount and MonthlyPayment, LoanOriginalAmount and LP_NetPrincipalLoss.


# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots-1}

# Scatter Plot for year vs CurrentDelinquencies color by IncomeRange.
# As per coach comments, I have changed the scale_color_brewe to use
# ordered color scheme and remove jitter.

# As per review comments, converted the scatter plot to box plot.
# Changed the y-scale to 5-20 to remove the data for CurrentDelinquencies
# which was causing overplotting.

#ggplot(aes(x = Year, y = CurrentDelinquencies,color=IncomeRange),
#       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_point() +
#  scale_color_brewer(palette='Oranges') +
#  facet_wrap(~ IncomeRange) +
#  theme(axis.text.x = element_text(angle = 90)) +
#  xlab('Year') +
#  ylab('CurrentDelinquencies') +
#  ggtitle('Number of delinquences per year By Income Range\n')

mp1 <- ggplot(aes(x = Year, y = CurrentDelinquencies),
              data = dfTemp2) +
  geom_boxplot(aes(fill = IncomeRange)) +
  scale_y_continuous(breaks=seq(5,20,5),limits=c(5,20)) +
  scale_fill_brewer(palette='Oranges') +
  facet_wrap(~ IncomeRange) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab('Year') +
  ylab('CurrentDelinquencies') +
  ggtitle('Number of delinquences per year By Income Range\n')

suppressWarnings(print(mp1))

# Sum of CurrentDelinquencies by Year
by(dfTemp$CurrentDelinquencies,dfTemp$Year,sum)

# Sum of CurrentDelinquencies by IncomeRange
by(dfTemp$CurrentDelinquencies,dfTemp$IncomeRange,sum)

```

In the above plot I have taken subset of deliquencies(5-20) and removed other data < 5 which was causing over plotting so plot looks clear.

From the plot we can see that in 2006 all of the deliquencies were from the borrowers with income range "Not displayed". Also there is a  drop of number of deliquencies in 2009. Now let's see the same based on Home owner type.

```{r echo=FALSE, Multivariate_Plots-2}

# Scatter Plot for year vs CurrentDelinquencies color by Borrower Type.
# As per coach comments, removed the jitter and introduced alpha=1/4
# for over plotting. 

#ggplot(aes(x = Year, y = CurrentDelinquencies,color=IsBorrowerHomeowner),
#       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_jitter(size = 2) +
#  scale_color_brewer(palette="Set1") +
#  theme_minimal() +
#  xlab('Year') +
#  ylab('CurrentDelinquencies') +
#  ggtitle('Number of delinquences per year By Borrower Type\n')

#ggplot(aes(x = Year, y = CurrentDelinquencies,
#           color=IsBorrowerHomeowner),
#       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_point(alpha=1/4) +
#  scale_color_brewer(palette="Set1") +
#  facet_wrap(~ IsBorrowerHomeowner) +
#  theme(axis.text.x = element_text(angle = 90)) +
#  xlab('Year') +
#  ylab('CurrentDelinquencies') +
#  ggtitle('Number of delinquences per year By Borrower Type\n')

# As per review comments, converted the above scatter plot to box plot.
# Changed the y-scale to 5-20 to remove the data for CurrentDelinquencies
# which was causing overplotting.

ggplot(aes(x = Year, y = CurrentDelinquencies),
       data = dfTemp2) +
  geom_boxplot(aes(fill = IsBorrowerHomeowner)) +
  scale_y_continuous(breaks=seq(5,20,5),limits=c(5,20)) +
  scale_fill_brewer(palette="OrRd") +
  facet_wrap(~ IsBorrowerHomeowner) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab('Year') +
  ylab('CurrentDelinquencies') +
  ggtitle('Number of delinquences per year By Borrower Type\n')


# Sum of CurrentDelinquencies by IsBorrowerHomeowner

by(dfTemp2$CurrentDelinquencies,dfTemp2$IsBorrowerHomeowner,sum)
by(dfTemp2$CurrentDelinquencies,dfTemp2$IsBorrowerHomeowner,summary)

```

In the above plot I have chosen the data with deiliquencies 5-20. We can see that Non Homeowner borrowers deliquencies are more than Home owner borrowers deliquencies in the years 2006 - 2008. Sum of deliquences are more for Non Homeowner borrowers.

Below is the sum of deliquencies for Homeowners and NonHomeowners based on data with deiliquencies 1-4

```{r echo=FALSE, Multivariate_Plots-2-1}
# Sum of CurrentDelinquencies 1-4 by IsBorrowerHomeowner
by(dfTemp1$CurrentDelinquencies,dfTemp1$IsBorrowerHomeowner,sum)
by(dfTemp1$CurrentDelinquencies,dfTemp1$IsBorrowerHomeowner,summary)

```

Below is the plot to see how Loan Amount varied based on Income Range per year.


```{r echo=FALSE, Multivariate_Plots-3}

# Scatter Plot for year vs LoanOriginalAmount color by IncomeRange.
# As per coach comments, reduced the jitter, used ordered color
# and faceted the plot to be more clear.

#ggplot(aes(x = Year, y = LoanOriginalAmount,color=IncomeRange),
#       data = LoanData) +
#  geom_jitter(alpha=0.5,position = position_jitter(width = .1)) +
#  scale_y_continuous(breaks = seq(0,40000, 5000),
#                     limits = c(0,40000)) +
#  scale_color_brewer(palette='OrRd') +
#  facet_wrap(~IncomeRange ) +
#  theme(axis.text.x = element_text(angle = 90)) +
#  xlab('Year') +
#  ylab('LoanOriginalAmount') +
#  ggtitle('Loans Issued per year By Income range\n')

# As per review comments, converted the above scatter plot to box plot.

ggplot(aes(x = Year, y = LoanOriginalAmount,color=IncomeRange),
       data = LoanData) +
  geom_boxplot(aes(fill = IncomeRange)) +
  scale_y_continuous(breaks = seq(0,25000, 5000),
                     limits = c(0,25000)) +
  scale_fill_brewer(palette='OrRd') +
  facet_wrap(~IncomeRange ) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab('Year') +
  ylab('LoanOriginalAmount') +
  ggtitle('Loans Issued per year By Income range\n')



```

Surprisingly all the loans from 2006 were of the Borrowers whose income range is "Not displayed". In the later years we can see a mix of borrowers with differnt Income ranges. In 2013 and 2014 there are many borrowers with income range above $100,000.


```{r echo=FALSE, Multivariate_Plots-4}

# Scatter PLot of CurrentDelinquencies vs LoanOriginalAmount 
# color by EmploymentStatus
# As per coach comments I have reduced the jitter size, introduced
# Alpha=1/10, used scale_color_brewer and increased the legend
# symbol size to make the clear plot.

#ggplot(aes(x = CurrentDelinquencies, 
#           y = LoanOriginalAmount, 
#           color = EmploymentStatus),
#       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_jitter(size = 5) +
#  scale_y_continuous(breaks = seq(0,40000, 5000),
#                     limits = c(0,40000)) +
#  theme_minimal() +
#  xlab('CurrentDelinquencies') +
#  ylab('LoanOriginalAmount') +
#  ggtitle('CurrentDelinquencies vs LoanOriginalAmount \n 
#  by EmploymentStatus\n')

ggplot(aes(x = CurrentDelinquencies, 
           y = LoanOriginalAmount, 
           color = EmploymentStatus),
       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
                       !is.na(CurrentDelinquencies))) +
  geom_jitter(alpha=1/10) +
  scale_y_continuous(breaks = seq(0,40000, 5000),
                     limits = c(0,40000)) +
  scale_color_brewer(palette='Set1') +
  theme_minimal() +
  guides(colour = guide_legend(override.aes = list(alpha = 1,size=5))) +
  xlab('CurrentDelinquencies') +
  ylab('LoanOriginalAmount') +
  ggtitle('CurrentDelinquencies vs LoanOriginalAmount 
          by EmploymentStatus\n')



# Sum of CurrentDelinquencies by EmploymentStatus

by(dfTemp$CurrentDelinquencies,dfTemp$EmploymentStatus,sum)

```

Above plot depicts that 85% of delinquent borrowers are Employed, Full-Time and Not available status. Borrowers with high Loan amounts (> $25000) have very fewer deliquencies compared to borrowers with Loan amounts < $1000.


```{r echo=FALSE, Multivariate_Plots-5}

# PLot of CurrentDelinquencies vs LoanOriginalAmount color by IncomeRange
# As per coach comments I have reduced the jitter size, introduced
# Alpha=1/5 for over plotting, used scale_color_brewer ,increased 
# the legend symbol size and feceted by IncomeRange to make the
# plot look nice and clear.

#ggplot(aes(x = CurrentDelinquencies, 
#           y = LoanOriginalAmount, 
#           color = IncomeRange),
#       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies))) +
#  geom_jitter(size = 8) +
#  scale_y_log10() +
#  scale_color_brewer(type = 'qual') +
#  theme_minimal() +
#  xlab('CurrentDelinquencies') +
#  ylab('LoanOriginalAmount') +
#  ggtitle('CurrentDelinquencies vs LoanOriginalAmount \n by IncomeRange\n')

mp5 <- ggplot(aes(x = CurrentDelinquencies, 
           y = LoanOriginalAmount, 
           color = IncomeRange),
       data = subset(LoanData,LoanData$CurrentDelinquencies > 0 & 
                       !is.na(CurrentDelinquencies))) +
  geom_jitter(alpha=1/5) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0,40,5),limits = c(0,40)) +
  scale_color_brewer(palette="BuGn") +
  guides(colour = guide_legend(override.aes = list(alpha = 1,size=5))) +
  theme_minimal() +
  facet_wrap(~ IncomeRange) +
  xlab('CurrentDelinquencies') +
  ylab('LoanOriginalAmount') +
  ggtitle('CurrentDelinquencies vs LoanOriginalAmount \n by IncomeRange\n')

suppressWarnings(print(mp5))

```

In the above plot, most of the borrowers whose Income Range is 'Not displayed' have deliquencies with Loan Amounts below $10,000. Deliquencies with high Loan Amounts are from borrowers with Income Range $100,000+.

```{r echo=FALSE, Multivariate_Plots-6}

# Plot of LoanOriginalAmount fill by IncomeRange and facet wrap 
# with Employment Status.
# As per coach comment I have added order color to income range.

ggplot(aes(x = LoanOriginalAmount, fill = IncomeRange),
       data = LoanData) +
  geom_histogram(binwidth=1000) +
  facet_wrap(~ EmploymentStatus) +
  scale_x_continuous(breaks = seq(0,20000, 5000),
                     limits = c(0,20000)) +
  scale_y_continuous(breaks = seq(0,8000, 2000),
                     limits = c(0,8000)) +
  scale_fill_brewer(palette="OrRd") +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("Count") +
  ggtitle("Loan counts for each of the Borrower's Employed Status\n")


```

Here in the above plot we can see Employed and Full-time borrowers with income ranges from $1-$24,999 to $100,000+ are issued more Loans. Part-time and Retired borrowers are issued very few number of loans.

```{r echo=FALSE, Multivariate_Plots-7}

# Create a new data frame 
# "LoanData.ld_by_CreditScoreRangeUpper_IsBorrowerHomeowner"
# the following variables in this order group by CreditScoreRangeUpper,
# IsBorrowerHomeowner and arrange by CreditScoreRangeUpper
# 1) CreditScoreRangeUpper
# 2) IsBorrowerHomeowner
# 3) mean_Loan_Amount
# 4) median_Loan_Amount
# 5) n
  
LoanData.ld_by_CreditScoreRU_IsBorrowerHomeowner <- LoanData %>%
  dplyr::group_by(CreditScoreRangeUpper,IsBorrowerHomeowner) %>%
  dplyr::summarise(mean_Loan_Amount=mean(LoanOriginalAmount),
                   median_Loan_Amount=median(as.numeric(LoanOriginalAmount)),
                   n =n()) %>%
  dplyr::ungroup() %>%
  arrange(CreditScoreRangeUpper)

#head(LoanData.ld_by_CreditScoreRangeUpper_IsBorrowerHomeowner)

# Line Plot between CreditScoreRangeUpper and mean_Loan_Amount 
# color with IsBorrowerHomeowner

mp7 <- ggplot(aes(x = CreditScoreRangeUpper, y = mean_Loan_Amount),
              data = LoanData.ld_by_CreditScoreRU_IsBorrowerHomeowner) +
  geom_line(aes(color=IsBorrowerHomeowner)) +
  xlab("\nCreditScoreRangeUpper") +
  ylab("Mean LoanorignalAmount\n") +
  ggtitle("Mean Loan Amount vs CreditScoreRangeUpper 
          by IsBorrowerHomeowner\n")

suppressWarnings(print(mp7))

```

In the above plot Mean Loan Amout increases as Credit Score increases for Non Home owners but for Home owners there is a fall  in the Mean Loan Amounts for the credit scores more than 800.

Now lets see the correlation coefficients for the below pairs using sampling data.
CreditScoreRangeUpper vs LoanOriginalAmount
LoanOriginalAmount vs MonthlyLoanPayment


```{r echo=FALSE, Multivariate_Plots-8}

theme_set(theme_minimal(10))

#Plots using ggpairs for CreditScoreRangeUpper vs LoanOriginalAmount 
#with sampling data of 10000

ld_subset1 <- LoanData[,c(9,14)]
#names(ld_subset1)
gg1 <- ggpairs(ld_subset1[sample.int(nrow(ld_subset1),10000), ])
suppressWarnings(print(gg1))

# Plots using ggpairs for LoanOriginalAmount vs MonthlyLoanPayment 
# with sampling data of 10000

ld_subset2 <- LoanData[,c(14,16)]
#names(ld_subset2)
ggpairs(ld_subset2[sample.int(nrow(ld_subset2),10000), ])



```

In the above plots, correlation co-efficient is high between LoanOriginalAmount and MonthlyLoanPayment.


```{r echo=FALSE, Multivariate_Plots-9}

# Scatter plot of CurrentDelinquencies vs CreditScoreRangeUpper
# facet wrap with IsBorrowerHomeowner
# As per coach commets, introduced Aplpha=1/5 for overplotting
# used scale_color_brewer with ordered color.

#ggplot(data = subset(LoanData,
#                     LoanData$CurrentDelinquencies > 0 & 
#                       !is.na(CurrentDelinquencies)), 
#       aes(x = CreditScoreRangeUpper, 
#           y = CurrentDelinquencies, 
#           color = IncomeRange)) +
#  geom_jitter(size=3) +
#  theme(axis.text.x = element_text(angle = 90)) +
#  facet_wrap(~ IsBorrowerHomeowner) +
#  scale_color_brewer(type = 'div') +
#  ggtitle("CreditScoreRangeUpper vs CurrentDelinquencies 
#           faceted by IsBorrowerHomeowner\n")


ggplot(data = subset(LoanData,
                     LoanData$CurrentDelinquencies > 0 & 
                       !is.na(CurrentDelinquencies)), 
       aes(x = CreditScoreRangeUpper, 
           y = CurrentDelinquencies, 
           color = IncomeRange)) +
  geom_jitter(alpha=1/5) +
  scale_color_brewer(type='seq') +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ IsBorrowerHomeowner) +
  guides(colour = guide_legend(override.aes = list(alpha = 1,size=5))) +
  ggtitle("CreditScoreRangeUpper vs CurrentDelinquencies 
           faceted by IsBorrowerHomeowner\n")


```

In the above plot most of the Borrowers with deliquencies if Homeowner or Non Homeowner, fall under credit score range of 400 to 700. Some Non Homeowner borrowers with high deliquencies have credit score of appx 500 and I think their Income Range is 'Not displayed'. In case of Non Homeowners all the borrowers with Income Range of 'Not displayed' are clustered between credit scores of 450-550. Borrowers with Income Ranges of $25,000-49,000 and $50,000-74,000 and $75,000-99,000 are clustered between 600 to 800 credit scores.

I want to build a linear model to predict the Loan Amount. I am using the varibales LoanOriginalAmount, MonthlyLoanPayment, EmploymentStatus, CreditScoreRangeUpper and IncomeRange

```{r echo=FALSE, Multivariate_Plots-10}
# Build a model using the varibales LoanOriginalAmount, MonthlyLoanPayment, 
# EmploymentStatus, CreditScoreRangeUpper and IncomeRange

m1 <- lm(I(log(LoanOriginalAmount)) ~ I(MonthlyLoanPayment^(1/3)), 
         data = LoanData)
m2 <- update(m1, ~ . + MonthlyLoanPayment)
m3 <- update(m2, ~ . + EmploymentStatus)
m4 <- update(m3, ~ . + CreditScoreRangeUpper)
m5 <- update(m4, ~ . + IncomeRange)
mtable(m1, m2, m3, m4, m5)

```

The variables in this model can account for 85% of variance in the Loan Amount issued to Borrower.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

After introducing "Year" and "CreditSCoreRangeUpper", it strengthened my other features of interest and I could see more details from the plots. 
Surprisingly in the Year 2006 most of the deliqeuncies were from borrowers whose Income range was "Not displayed". May it was because Prosper formed that year and guide lines might not be stricter. In 2009 number of deliquencies dropped significantly as the guide lines were made stricter. Employed and FullTime borrowers had more delinquencies as they have more loans compared to other borrowers. Non Homeowners have more deliquencies. Borrowers with high credit scores have loans with lower interest rates.


### Were there any interesting or surprising interactions between features?

I did a further investigation by taking the mean of "LoanOrignalAmount" and plotting against "CreditSCoreRangeUpper". I found that for Non Homeowners Loan Amount increased by credit score but for Homeowners I see a drop in the LoanAmount for the credit scores more than 800.


### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

I created a Model using the varibales log10 of LoanOriginalAmount, cube root of MonthlyLoanPayment, EmploymentStatus, CreditScoreRangeUpper and IncomeRange and found that this model can account for 85% of variance in the Loan Amount issued to Borrower.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

# Box Plot of LP_NetPrincipalLoss per year based on Income Range.
# As per coach comments I have faceted the box plots based on
# IncomeRange so that it is more readable.


ggplot(aes(x = Year, 
           y = LP_NetPrincipalLoss ), 
       data = subset(LoanData,
                     LoanData$LP_NetPrincipalLoss > 0)) +
  geom_boxplot(aes(fill=IncomeRange)) + 
  facet_wrap(~IncomeRange) +
  scale_fill_brewer(type='seq') +
  ylab('LP_NetPrincipalLoss') +
  xlab('year') +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle('Net principal Loss per year\n')


```

### Description One

In Bi-Variate section I had Observed how the Net Principal Loss varied each year. Now here I want to see how the Net Principal Loss varied each year among the borrowers with different Income ranges. So In the above plot I have used different colors for different Income Ranges. It is clear that Median Net principal loss is high in the year 2007 and was low in 2009 when prosper implemented strict guide lines. Then it increased again from  2011 to 2013 for all Incomeranges except for borrowers with Incomerange of $0 and 'Not displayed'. For all the years, Income Range of $100,000+ has higher median of Net principal Loss.


### Plot Two
```{r echo=FALSE, Plot_Two}

# Scatter plot of CreditScoreRangeUpper vs CurrentDelinquencies
# facet wrap with EmploymentStatus
# to reduce overplotting or remove outliers use
# CurrentDelinquencies > 1. Also reduce the y-axis scale to 40.
# As per reviewer comments removed the CreditScoreRangeUpper values
# around 0 and changed the scale to 400-800.

fp2 <- ggplot(data = subset(LoanData,
                            LoanData$CurrentDelinquencies > 1 &
                              !is.na(CurrentDelinquencies)),
              aes(x = CreditScoreRangeUpper,
                  y = CurrentDelinquencies,
                  color = IncomeRange)) +
  geom_jitter(alpha=1/5) +
  scale_y_continuous(breaks = seq(5,50,5),limits = c(5,50)) +
  scale_x_continuous(breaks = seq(400,800,100),limits=c(400,800)) +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ EmploymentStatus) +
  guides(colour = guide_legend(override.aes = list(alpha = 1,size=5))) +
  scale_color_brewer(palette="OrRd") +
  ggtitle("CreditScoreRangeUpper vs CurrentDelinquencies 
           faceted by EmploymentStatus\n")

suppressWarnings(print(fp2))

```

### Description Two

In Bi-variate section I saw that the  the median of delinquencies is high at credit score of 459, 479, 499, 519 and dereased by increase in credit score upto 699. In Multi-Variate section I saw some Non Homeowner borrowers with high deliquencies have credit score of appx 500 and I think their Income Range is 'Not displayed'. In case of Non Homeowners all the borrowers with Income Range of 'Not displayed' are clustered between credit scores of 450-550. Now I want to see how the number of deliquencies varies with credit score based on borrowers employment status.

In the above plot I am considering Deliquencies which are greater than 4 similar to bi-variate section (as there as manyy entries with 1 and causing overplotting at the bottom of the plot.) and reduced the y-axis scale to 50 so that plots are clear.

For all borrowers of various Income Ranges with Deliquencies, Credit Score range is between ~500-800. I can see there are many Deliquenicies for the Borrowers whose Employment Status is 'Employed' or 'Full-time' and Income Range varies. Also Borrowers whose Employment Status is 'Not available' or not specified have many delinquencies with IncomeRange of 'Not displayed'.  Deliquencies > 20 are distributed among Borrowers whose Employment Status is not specified,'Not available' and 'Full-time' with various Income Ranges.'Not Employed' and 'Retired' Borrowers have least Deliquencies. There are some Borrowers with Delinquencies whose credit score is way below 50. They belong to Employment Status not specified category. It would be interesting to know who they are.


### Plot Three
```{r echo=FALSE, Plot_Three}

# Scatter PLot of BorrowerRate vs LoanorignalAmount, 
# color by CreditScoreRangeUpper
# and faceted with year. Limited the CreditScoreRangeUpper from 600 to 850 
# to show the colors clearly. Used suppressWarnings and print to suppress
# the warnings generated by the plot.

fp3 <- ggplot(data = LoanData, 
       aes(x = BorrowerRate,
           y = LoanOriginalAmount)) +
  geom_jitter(alpha=0.5,aes(color = CreditScoreRangeUpper),
              size=2) +
  scale_x_continuous(breaks = seq(0.0,0.3, 0.1),
                     limits = c(0.0,0.3)) +
  scale_color_gradient(limits=c(600,850),
                       high="green",
                       low="blue") +
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~ Year) +
  xlab("Borrower Rate") +
  ylab("Loan Amount") +
  ggtitle("Loan Amount vs Borrower Rate by Credit Score each year\n")

suppressWarnings(print(fp3))

```

### Description Three

In the Uni-variate section I saw a spike in count for the borrower rate of 0.3177. On further analysis in Bi-variate section I have observed Many borrowers with 'Employed' status with various Income Ranges have this rate.In this section I want to see how BorrowerRate for Loan Amounts varies by CreditScoreRangeUpper for each year.

I can see from above plot, borrowers with high Upper Range Creditscore are in the left of the plot and they generally have low interest rates. From 2010 to 2014 we can see Low Upper Range credit score borrowers have high interest rates. I think this is because of making the guide lines strict in 2009. In 2006,2007,2008 I see borrowers with credit score in the range 600-700 have many loans. In 2009 the number of loans for them decreased and gradually increased again from 2011 to 2014. 

------

# Reflection

Prosper Loan Data set contains information about ~114,000 loans issues to all kinds of borrowers from 2006 to 2014. It has 81 variables out of which I have chosen 21 variables to explore the data. It took some time to understand the varibales and select the from the list of 81 for futher exploration. I had to spend time to understand how Prosper works so that it will be easy for me to explore the varibles I have chosen.

From my analysis it seems like Propser struggled in the intial years until 2008 with many deliquencies, no strict rules etc. As we can see there was no proper information like Income Range, Employment Status about the borrowers. In 2009 once Prosper was restructured and made the rules stricter, I can see the deliquencies dropped significantly.

I would like to explore more on how Interest and fees are varying among different Loans, how the loans are distributed accross different states. Also I want to learn more about the models and create more models on this data using the other features in the data for example model loans likely to be deliquent based on credit scores.

# References

- Common function in R: http://discussions.udacity.com/t/common-function-in-r/10404/2
- Reshape: http://www.statmethods.net/management/reshape.html
- Date Formats in R: http://www.r-bloggers.com/date-formats-in-r/
- Do more with dates and times in R with lubridate: http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html
- Introduction of dplyr: dplyr Tutorial Part 1 : http://www.r-bloggers.com/hadley-wickhams-dplyr-tutorial-at-user-2014-part-1/
- Types of smoothers in ggplot2: http://www.ats.ucla.edu/stat/r/faq/smooths.htm
- An Introduction to reshape2 by Sean Anderson: http://seananderson.ca/2013/10/19/reshape.html
- Converting Between Long and Wide Format: http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/
- Melt Data Frames: http://www.r-bloggers.com/melt/
- Cut Function: http://www.r-bloggers.com/r-function-of-the-day-cut-2/
- Bayesian Statistics and Marketing: http://www.perossi.org/home/bsm-1
- QuickR's Graphical Parameters: http://www.statmethods.net/advgraphs/parameters.html
- facet_wrap: http://docs.ggplot2.org/0.9.3.1/facet_wrap.html
- Log Transformations: http://www.r-statistics.com/2013/05/log-transformations-for-skewed-and-wide-distributions-from-practical-data-science-with-r/
- Basic Structure of a Function: https://www.youtube.com/watch?v=Z1wB1rHAYzQ&list=PLOU2XLYxmsIK9qQfztXeybpHvru-TrqAP
- Defining a New Transformation in ggplot2 and the Scales Package: http://blog.ggplot2.org/post/25938265813/defining-a-new-transformation-for-ggplot2-scales
- ggplot2: axis manipulation and themes: http://rstudio-pubs-static.s3.amazonaws.com/3364_d1a578f521174152b46b19d0c83cbe7e.html
- An Introduction to corrplot package: http://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html
- R Sub function: http://www.endmemo.com/program/R/sub.php
- ggplot2 - colors: http://docs.ggplot2.org/0.9.3.1/scale_gradient.html
- Beautiful plotting in R: A ggplot2 cheatsheet: http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/
- Prosper: https://www.prosper.com/invest/marketplace-performance/
- Prosper: http://en.wikipedia.org/wiki/Prosper_Marketplace
- Prosper Ratings: https://www.prosper.com/invest/how-to-invest/prosper-ratings/
- Prosper: How it works: https://www.prosper.com/welcome/how-it-works/
